---
title: "Using Cartographer for Interactive Maps in R"
author: "Lincoln Mullen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

```{r, echo=FALSE}
library(knitr)
opts_chunk$set(fig.width = 7.2, fig.height = 4.25)
```

The [cartographer package](https://github.com/lmullen/cartographer) for R lets you create interactive, exploratory maps inside R. ([Installation instructions](https://github.com/lmullen/cartographer).) These maps are intended for use in data exploration. Cartographer provides the ability to plot points and scale them, to draw GeoJSON or TopoJSON polygons, and to use map tiles from popular providers.

Cartographer is an interface to the [d3-carto-map](https://github.com/emeeks/d3-carto-map) JavaScript library written by [Elijah Meeks](https://twitter.com/elijah_meeks), which in turn provides additional mapping functions to the [D3.js](http://d3js.org/) JavaScript library by [Mike Bostock](http://bost.ocks.org/mike/). Thanks to the [htmlwidgets](http://www.htmlwidgets.org/) package, the maps it creates can be embedded in [R Markdown](http://rmarkdown.rstudio.com/) documents or used at the R console.

To demonstrate the uses of the cartographer package, we will use the spatial data provided by the [historydata](https://github.com/ropensci/historydata) package and [dplyr](http://cran.rstudio.org/web/packages/dplyr/) for data manipulation.

```{r, message=FALSE}
library(cartographer)
library(historydata)
library(dplyr)
```

## Basic Use

A cartographer map is composed using functions that add layers; these layers are glued together by the [magrittr](https://github.com/smbache/magrittr) pipe operator, `%>%`.

Every map must start with the `cartographer()` function. This function controls options for the entire map. But the map won't appear without at least one layer. We'll add some tiles with `tile_layer()`.

```{r}
cartographer() %>%
  tile_layer()
```

## Tile layers

Tile provide a zoomable base layer underneath the data that you wish to plot. To change the tiles you need to specify a tile provider and an identifier for the tiles. Cartographer lets you use tiles from [Stamen](http://maps.stamen.com/#toner/12/37.7706/-122.3782) and [Mapbox](http://a.tiles.mapbox.com/v3/mapbox/maps.html). Below are a sample set of tiles from each.

Stamen's "Toner Background" tiles:

```{r}
cartographer() %>%
  tile_layer(provider = "stamen", path = "toner-background")
```

Mapbox's "World Bright" tiles:

```{r}
cartographer() %>%
  tile_layer(provider = "mapbox", path = "mapbox.world-bright")
```

Mapbox's "Geography Class" tiles:

```{r}
cartographer() %>%
  tile_layer(provider = "mapbox", path = "mapbox.geography-class")
```

Mapbox's "Natural Earth Hypso/Bathy":

```{r}
cartographer() %>%
  tile_layer(provider = "mapbox", path = "mapbox.natural-earth-hypso-bathy")
```

## Points layers

Points can be displayed on the map using a data frame with columns for latitude and longitude (in decimal degrees) and other columns with additional data. We will create a map of the Catholic dioceses in North America using the `catholic_dioceses` data set in the historydata package.

```{r}
data(catholic_dioceses)
catholic_dioceses
```

### Basic Points Layer

The data frame should be passed to the `points_layer()` function. You must identify the columns containing longitude and latitude with the `x =` and `y =` parameters, respectively, unless those columns happen to be titled `long` and `lat`. 

```{r}
cartographer(region = "United States") %>%
  tile_layer() %>%
  points_layer(data = catholic_dioceses, x = "long", y = "lat")
```

### Multiple Points Layers

One of the advantages of cartographer is the ability to add multiple point layers. We will separate out the archdioceses from the regular dioceses, then create separte layers for each. We can control the `size`, `color`, and `opacity` of these layers, along with the `label` in the layer selector, with arguments to `point_layer()`.

```{r}
archdioceses <- catholic_dioceses %>% filter(event == "metropolitan")
dioceses     <- catholic_dioceses %>% filter(event == "erected")

cartographer(region = "united states") %>%
  tile_layer() %>%
  points_layer(data = dioceses, color = "green", size = 4,
                label = "Dioceses") %>%
  points_layer(data = archdioceses, color = "purple", size = 8, opacity = 1,
                label = "Archdioceses")
```

### Clustered Points

When you have large numbers of points, it may make sense to cluster the points. The points will cluster together when you zoom out, and disaggregate when you zoom in. 

```{r}
cartographer(region = "United States") %>%
  tile_layer() %>%
  points_layer(data = catholic_dioceses, cluster = TRUE)
```

### Scaled points

If you have additional numeric data besides latitude and longitude in your data frame, then you can make the area of the points proportional to the data using. The argument `radius_field =` identifies which column should be used. The `size =` argument will control the maximum size of the points. The scale begins at zero. Here we use the `paulist_missions` data from the historydata package. It has two columns, `confessions` and `converts` which indicate the relative success that the Catholic missionary order had at each mission. We will add a layer for each but only display the layer for confessions at the start.

```{r}
data(paulist_missions)

cartographer(region = "United States") %>%
  tile_layer() %>%
  points_layer(data = paulist_missions, radius_field = "confessions", size = 12,
               label = "Confessions") %>%
  points_layer(data = paulist_missions, radius_field = "converts", size = 12,
               color = "green", label = "Converts", visible = FALSE)
```

If you wish you can modify the [D3 quantitative scale](https://github.com/mbostock/d3/wiki/Quantitative-Scales) used to scale the points by passing a different JavaScript function to the `radius_func =` parameter.

## TopoJSON and GeoJSON Layers

Cartographer can display [TopoJSON](https://github.com/mbostock/topojson) and [GeoJSON](http://geojson.org/) polygons using `topojson_layer()` and `geojson_layer()`. The data for these layers can be passed as a string to the `data =` parameter. But usually it is more convenient to pass the path to a TopoJSON or GeoJSON file using the `file = ` parameter. The `fill =` and `stroke =` colors can be set, as can the `opacity =`.

### GeoJSON

Here is a GeoJSON layer of land polygons using a sample file.

```{r}
land <- system.file("extdata", "land.geojson", package = "cartographer")
cartographer() %>%
  geojson_layer(file = land, label = "Land")
```

### TopoJSON

When creating a TopoJSON layer, you must specify which feature to map using the `feature =` parameter. Here is TopoJSON file of state and territory boundaries for the United States of America in June 1853. The feature name is `us`.

```{r}
us <- system.file("extdata", "us1853.topojson", package = "cartographer")
cartographer(region = "United States") %>%
  topojson_layer(file = us, feature = "us", label = "US 1853")
```

## Options

### General

General options are controlled in the `cartographer()` function.

You can set the initial zoom of a map by using the `region =` argument. The region can be a two-letter ISO country code, the name of a continent, or "United States" for the continental USA.

Continent:

```{r}
cartographer(region = "Africa") %>%
  tile_layer()
```

Country:

```{r}
cartographer(region = "de") %>%
  tile_layer()
```

Continental United States of America:

```{r}
cartographer(region = "United States") %>%
  tile_layer()
```

### Layer specific

You can include as many layers as you like. You may wish to turn off some layers by default. Any layer can turned off by setting `visible = FALSE` as an argument to the layer function. These layers can later be toggled on with the layer selector.

If point or TopoJSON/GeoJSON layers contain additional properties, such as the names of regions or data associated with a point, a popup will appear with that information if they are clicked. By default, points can be clicked and polygons cannot, but this can be controlled with the `clickable =` parameter to `topojson_layer()`, `geojson_layer()`, and `points_layer()`.

Layer labels in the layer selector can be set with the `label =` parameter to a layer function.
